Contrastive Optimization -- Preliminaries

### 2024-10-28 paper/main.tex
bp-free highres-cams, optimized finetuning
fun fact: till now, the last layer was *always* just the same as when it was initialized randomly
added experiments to establish motivation for contrastive approach, added models, made minor bugfixes

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{bbm}
\usepackage{bm}

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class. The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
\begin{gather}
	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\}^{|C|-1}_{c_{t'} \in c - c_t}
\end{gather}
This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.

@@

@@

As discussed in the previous section, the distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

@@

@@

\begin{theorem}\label{ablatedce}
	We can integrate background invariance to the definition of cross-entropy using the following formulation:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
\end{theorem}
\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from \ref{classprobdiff}, we have:
	\begin{align}
		d &= [s_i - s_c]_i \\
		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
	\end{align}
	As an aside: we note that when $i = c$, the contrastive map is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps. \newline \\
	Using our result from \ref{classprobdiff}, we have:
	\begin{gather}
		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
	\end{gather}
	For a target class $c$, cross-entropy loss is defined as:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right)
	\end{gather}
	We can break down each contrastive map to foreground and background components:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} - \sum (1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)} \right)
	\end{gather}
	The current loss formulation motivates improving contrast using either the foreground or the background. We penalize contrast on the background, motivating learning invariance to noise in the form of the background.
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
	This gives us our ablated cross-entropy formulation.
\end{proof}

@@

@@

The Oxford-IIIT Pets dataset is an image dataset with a classification objective alongside high-quality segmentation maps. We train the model on the binary classification case (image containing a dog or cat) as well as the fine-grained, breed specific 37 class split. The architecture used for training was ResNet-50, initialized with random parameters and three key modifications:

@@

\subsection{Training Invariance to Spurious Correlation}

One significant concern uncovered both theoretically from \ref{ablatedce} and empirically from \ref{contr-cam} was that the standard classification pipeline motivates learning spurious correlations between the model, provided the spurious information is able to improve accuracy. This motivates learning distribution-specific correlations, which worsens it's ability to generalize. As an example

\begin{figure}[H]
	\centering
	\includegraphics[width=.3\textwidth]{img/default_cam.png}
	\hspace{1em}
	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
	\hspace{1em}
	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
	\caption{Left: contrastive activation maps from the default training run. Middle: contrastive activations maps from the Ablated CE+KLD training run. Right: contrastive activation maps from the Ablated CE training run.}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=.3\textwidth]{img/default_cam.png}
	\hspace{1em}
	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
	\hspace{1em}
	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
	\caption{Left: contrastive activation maps from the default training run. Middle: contrastive activations maps from the Ablated CE+KLD training run. Right: contrastive activation maps from the Ablated CE training run.}
\end{figure}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{subfig}
\usepackage{float}
\usepackage{bbm}
\usepackage{bm}

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} ~~\text{ where }~~ \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c')} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information in the form of the activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class. The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
\begin{gather}
	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c - c_t}
\end{gather}
This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.

@@

\subsection{Observation: Cross-Entropy Motivates Learning from Spurious Correlations}

Using contrastive activation maps, we can evaluate the attention maps generated by a \underline{well-trained} model (ImageNet weights, fine-tuned to $99.9\%$ training accuracy, $99.6\%$ validation accuracy) on Oxford-IIIT Pets:

\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{img/default_cams.png}
\end{figure}

In this random sample, we observe that the regions within each of the images that influence the final prediction are misaligned to the classification target. Particularly, the top-left corner of the final two images contributed significantly to the final prediction, despite not containing the classification target.

Interestingly, this dependence holds even when we average influence plots generated from inference across the dataset:
\begin{figure}[h!]
	\centering
	\subfloat[Pre-Trained on Train]{\includegraphics[width=.24\textwidth]{img/default_pretrained_finetuned_best_train_influence.png}}
	\subfloat[Pre-Trained on Valid]{\includegraphics[width=.24\textwidth]{img/default_pretrained_finetuned_best_valid_influence.png}}
	\subfloat[Random Init on Valid]{\includegraphics[width=.24\textwidth]{img/default_best_valid_influence.png}}
	\subfloat[Random Init on Train]{\includegraphics[width=.24\textwidth]{img/default_best_train_influence.png}}
\end{figure}

Further, we can use the segmentation trimaps from the Oxford-IIIT Pets dataset as foreground and background delimiters to bifurcate the regions that influence final prediction logits. On average, we observe the following:

\begin{table}[h]
	\centering
	\begin{tabular}{c|cc|cc}
		\toprule
		\bf Average & \multicolumn{2}{c}{\bf Validation} & \multicolumn{2}{c}{\bf Training} \\
		\bf Pixel Influence & \textbf{Random} & \textbf{Pre-Trained} & \textbf{Random} & \textbf{Pre-Trained} \\
		\midrule
		Foreground & 6.834 & \bf 2.622 & 6.673 & \bf 2.579 \\
		Background & \bf 8.934 & 1.347 & \bf 8.738 & 1.332 \\
		\midrule
		Net & 15.769 & 3.970 & 15.410 & 3.912 \\
		\bottomrule
	\end{tabular}
\end{table}

Crucially, we observe that weights optimized from random initialization leverage more of the background than the actual target to inform the final prediction. In training classification models, an implicit assumption is that the best approximator of the class label is the target itself, which experimentally are observed to not hold. Despite fine-tuning on pre-trained weights mitigating the problem, it still learnt a feature that can be exploited adversarially.

This is especially concerning for cases wherein the target is far away from the camera, in which case the emphasis is objectively placed on learning the best spurious surrogate to the actual target, rather than obtaining an approximation of the target's feature representation using the fewer available pixels.

% TODO: adversarial ablation

@@

The primary distinction between the contrastive optimization objective over standard risk is that it preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

@@

% TODO: formal analysis -- empirical performance test on scale augmentation + norm-change in feature representation.

In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target also motivates learning scale-robust representations of the target.

@@

\begin{theorem}\label{ablatedce}
	We can integrate backgroud invariance to the definition of cross-entropy using the following formulation:
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
\end{theorem}
\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from \ref{classprobdiff}, we have:
	\begin{align}
		d &= [s_i - s_c]_i \\
		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
	\end{align}
	As an aside: we note that when $i = c$, the contrastive map obtained is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps. \newline \\
	Using our result from \ref{classprobdiff}, we have:
	\begin{gather}
		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
	\end{gather}
	For a target class $c$, cross-entropy loss is defined as:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right)
	\end{gather}
	We can break down each contrastive map to foreground and background components:
	\begin{gather}
		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} - \sum (1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)} \right)
	\end{gather}
	The current loss formulation motivates improving contrast using either the foreground or the background. We penalize contrast on the background, motivating learning invariance to noise in the form of the background.
	\begin{gather}
		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{gather}
	This gives us our ablated cross-entropy formulation.
\end{proof}

@@

\subsubsection{ACE-Divergence Formulation}

An empirical observation from the behavior of cross-entropy loss is that it displays a tendency to generate contrast only to regions where feature differences are present within the dataset, which places an assumption for the same set of differenciating features being present across the underlying task. In addition, the peaky nature of the generated contrastive maps also encourages vulnerability to adversarial pertubation.

We therefore combined the divergence based loss with ablated cross-entropy. The ablated cross-entropy term ensures accurate predictions using only the region where the classification target is present, and the divergence term mitigates the dataset-specific contrastive setting, motivating effectively learning an implicit segmentation map.

\begin{gather}
	\begin{split}
		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)})) \\
		&- \lambda_4 \log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
	\end{split}
\end{gather}

For this setting, $\bm{\lambda} = \{1, 50, 1, 10^{-3}\}$ was set for $i = \{1, \ldots, 4\}$ respectively. As in the previous setting, these were identified empirically, without rigorous optimization.

@@

% \begin{figure}
% 	\centering
% 	\includegraphics[width=\textwidth]{img/cams.pdf}
% \end{figure}

The Oxford-IIIT Pets dataset is an image dataset with a classification objective alongside high-quality segmentation maps. We train the model on the binary classification case (image containing a dog or cat) as well as the fine-grained, breed specific 37 class split. The architecture used for training was ResNet-50, initialized either with random parameters or ImageNet, with three key modifications:

@@

@@

\begin{figure}[H]
	\centering
	\subfloat[Cross-Entropy (CE)]{\includegraphics[width=.3\textwidth]{img/default_cam.png}}
	\hspace{1em}
	\subfloat[Ablated CE]{\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}}
	\hspace{1em}
	\subfloat[Ablated CE+KLD]{\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}}
	\caption{Contrastive Class Activation Maps from each training setting.}
\end{figure}
